An Ordered Logic for Category Theory

### 2022-09-07 paper/paper.tex
related work

\paragraph{Calculi for 2-dimensional category theory}

@@

Our type-theoretic syntax for virtual double categories is intended to
be a formal notation for the manipulation of profunctors. As such it
has similarities to informal syntax for manipulating profunctors
involving ends and coends, such as the coend calculus of
\citet{coend-calculus}.

@@

Double-cat-like syntax:
Gradual Type Theory, System P of parametricity, cubical stuff?

\subsection{Directed Type Theory}

Directed Type theory

Fibrational approaches to logical relations

>>>>

\paragraph{Calculi for formal category theory}

In their book on $\infty$-category theory, Riehl and Verity use the
formal language of virtual equipments to prove results valid for
$\infty$-categories without concrete manipulation of model
categories. They formalize this language using Makkai's FOLDS
framework of first-order logic with dependent sorts. The approach is
similar in spirit to our \ohol{}, and has the same models. The main
difference is that their theory is built on top of first-order logic,
whereas ours is built on a custom-made variant of higher-order logic
over a (unary) simple type theory. For this reason, statements in
their theory are quite complex as they have to be encoded into
first-order logic, whereas we argue our type-theoretic approach
formalizes informal arguments much more directly, as shown in
Section~\ref{sec:examples}.
%
An informal syntax for manipulating profunctors involving ends and
coends, called the coend calculus is closely related, an extension of
\ohol{} to treat profunctors of many variables of different variances
should provide a formal treatment for the coend calculus
\citet{coend-calculus}.

Another approach to formal category theory is to develop a
\emph{directed dependent type theory}, a type theory where the
identity type is interpreted as morphisms in a (possibly
$\infty$-)category. Research on directed dependent type theories is
ongoing \cite{riehl-shulman,north}. Directed dependent type theories
should generalize \ohol{} in the same way that dependent type theory
generalizes higher order logic: internally there should be a type that
models the category of sets, and terms of this type should model the
profunctors of \ohol{}. As a foundation for a domain-specific proof
assistant for category theory, \ohol{} may provide a foundation for a
proof assistant in the style of Isabelle/HOL \cite{isabelle-hol},
whereas directed type theory would provide a foundation for a proof
assistant in the style of Agda or Coq \cite{agda,coq}.

@@

@@
